Forecasting Gamma-Ray Bursts using Gravitational Waves

### 2018-08-16 Paper/Manuscript.tex
Final review, modulo figures and TT gauge sentence

\begin{abstract}
We explore the intriguing possibility of employing future ground-based gravitational-wave interferometers to detect the inspiral of binary neutron stars sufficiently
early to alert electromagnetic observatories so that a gamma-ray burst (GRB) can be observed in its entirety from its very beginning.
We quantify the ability to predict a GRB by computing the time a binary neutron star (BNS) system takes to inspiral from its moment of detection to its final merger. We define the moment of detection to be the instant at which the interferometer network accumulates a signal-to-noise ratio of 15. %for the BNS inspiral.
For our computations, %of advance warning times, 
we specifically consider BNS systems at luminosity distances of (i) $D\le200\,$Mpc in the three-interferometer Advanced-LIGO-Virgo network of 2020, and (ii) $D \le 1000\,$Mpc in the Einstein Telescope's B and C configurations. 
In the case of Advanced LIGO-Virgo we find that we may at best get a few minutes of warning time, thus we expect no forecast of GRBs in the 2020s. 
On the other hand, Einstein Telescope will provide us with advance warning times of more than five hours for $D \le 100\,$Mpc.
Taking one hour as a benchmark advance warning time, we obtain a corresponding horizon
distance of roughly 600 Mpc for the Einstein Telescope C configuration.
%, more specifically, greater than an hour
%for BNS inspirals closer than roughly 600\,Mpc. Taking this luminosity distance to be our horizon distance, 
Using current BNS merger event rates within this volume, we show that Einstein C will forecast $\gtrsim \ord(10^2)$ GRBs in the 2030s. %with Einstein C. %with the C configuration of Einstein Telescope.
We reapply our warning-time computation to binary black hole - neutron star inspirals and find that we expect 1 to 3 tidal disruption events 
to be forecast by the same detector.
This article is intended as a pedagogical introduction to gravitational-wave astronomy written at a level accessible to
Ph.D. students, advanced undergraduates, and colleagues in astronomy and/or astrophysics who wish to learn more about the underlying physics.
Though many of our results may be known to the experts, they might nonetheless find this article motivating and exciting.
\end{abstract}
\maketitle

\section{Introduction}\label{Sec:Intro}
The 6.95\footnote{There have thus far been six detections with $5\sigma$ statistical significance and one additional event, LVT151012 \cite{TheLIGOScientific:2016pea}, with $2\sigma \simeq 95\%$ significance hence 6.95 overall.} 
gravitational-wave merger events detected by the Advanced Ligo-Virgo network have firmly established gravitational-wave astronomy
as an observational science \cite{TheLIGOScientific:2016pea, Abbott:2017gyy, Abbott:2017oio, Abbott:2017vtc, GW170817}. 
Though the first event of 14 September 2015 (GW150914 \cite{GW150914}) will always be the ``poster-child'' of this field,
the icing on the cake was the 17 August 2017 (GW170817 \cite{GW170817}) event involving the inspiral and merger of a binary neutron star system 
promptly followed, 1.7 seconds later, by a gamma-ray burst (GRB170817A) detected by the Fermi Gamma-ray Burst Monitor \cite{Fermi} and by the International Gamma-Ray Astrophysics Laboratory \cite{Svinkin, Savchenko:2017ffs}.
The Advanced-LIGO-Virgo network's initial source localization to within $ 31\,\text{degree}^2$ in the Southern skies
enabled astronomers to locate the electromagnetic counterpart in the galaxy NGC 4993, 
first picked up by the One-Meter, Two Hemisphere team less than 11 hours after the merger \cite{Coulter2017, Coulter:2017wya}.
This subsequently launched a massive campaign of multi-messenger astronomy across the entire electromagnetic (EM) spectrum which is still ongoing
more than half a year after the initial GRB \cite{GBM:2017lvd}.

GW170817 was initially identified by the LIGO-Hanford interferometer (H1) 
using a template bank of gravitational waveforms computed within the framework of post-Newtonian theory \cite{Buonanno:2009zt, Blanchet_LRR}.
The inspiral swept across the Advanced Ligo-Virgo (ALV) network's bandwidth from approximately $ 30\,$Hz to about $ 2\,$kHz in 57 seconds,
executing nearly $ 3000$ gravitational wave cycles and accumulating a network signal-to-noise ratio of 32.4 with a false-alarm rate of one per $8.0\times 10^4$ years \cite{GW170817}. 
The total mass of the system was inferred to be $2.74^{+0.04}_{-0.01} M_\odot$ with component masses in the range $1.17-1.60 M_\odot$, consistent
with neutron stars (assuming low spin priors, see Sec.~\ref{sec:idealizations}).
Moreover, both the gravitational wave (GW) and EM observations were consistent with a source location in the galaxy in NGC4993 
at a luminosity distance of $\sim\!\! 40\,$Mpc,
and supported the hypothesis that GW170817 resulted from the inspiral and merger of two neutron stars, 
causing a prompt short-hard gamma-ray burst (GRB170817A) as first suggested by Ref.~\cite{Eichler:1989ve}.

A week after GW170817, the ALV network was taken offline for upgrades for the third observation run (O3) 
scheduled to start in Autumn 2018 with roughly twice the sensitivity of O2.
In addition, the Japanese cryogenic interferometer KAGRA (\cite{KAGRA, KAGRA2}) will start its test runs in 2018-2019 and reach its design sensitivity circa 2021-2022 \cite{Akutsu:2017thy, Aasi:2013wya}. By 2025, LIGO-India detector will join the global interferometer network hence improving both the overall sensitivity and the sky-localization capabilities \cite{Aasi:2013wya}.
Around this time, the construction of the first third-generation ground-based interferometer, Einstein Telescope, will begin in Europe \cite{ET_doc}. 
Meanwhile, the US is currently considering a mid-2020s update to LIGO called Voyager \cite{LIGO_Voy} and the construction of a more ambitious 40\,km long interferometer dubbed The Cosmic Explorer \cite{CE}.
All of these near-future additions to the GW-detector network beg a simple question: can we detect the inspiral of a binary neutron star system
early enough to witness a gamma-ray burst as it happens? In other words, can we forecast GRBs using GW detections?

The short answer is ``yes''. The long answer is still ``yes'', but depends on many factors such as source distance and sky position, detector sensitivity and orientation. One also needs to account for the fact that GRBs are thought to be collimated emissions with narrow outburst angles \cite{Kumar:2014upa}.
As such, most GRBs that occur in the universe are electromagnetically undetectable. Fortunately, this was not the case with GRB170817A, 
but even for a reasonable GRB jet opening angle of $\sim 10^\circ$, we expect to observe $\sim 10\%$ of neutron star mergers as GRBs \cite{Patricelli:2016bkt}. 
However, it is possible that some of the near-Earth pointing short hard GRBs may be observed as long-duration GRBs or X-ray flashes thus increasing this percentage \cite{Bucciantini:2011kx}.

@@

Here, we are interested  in exploring the predictive power of near-future ground-based GW detectors. 
%thus leave the task of reducing the latency to the experts.
More specifically, we wish to find out how much early warning the GW detectors can provide us before the merger/GRB.
Ref.~\cite{Cannon:2011vi} approached this from an algorithmic perspective by developing computationally inexpensive methods to detect inspiral
signals in GW data to provide early-warning triggers \emph{before} the merger (also see Ref.~\cite{Knowles:2018hqq} for a very recent improvement to what is currently being
used in the ALV pipeline).
We instead focus on the forecasting capabilities of the ground-based interferometers at their \emph{full design} sensitivity by computing the time interval
between the merger and the instant of detection defined in terms of a certain detection criterion which we describe below.
To this end, we consider the ALV network of 2020s composed of three L-shaped interferometers operating at their design sensitivities
and Einstein Telescope of 2030s made up of V-shaped interferometers.
Ref.~\cite{Nissanke:2012dj} investigated the capabilities of three to five-interferometer detector networks of 2020s in terms of their horizon distances
to inspiralling BNSs and their sky-localization of these sources. Our work here is somewhat complementary in the sense that we are mostly concerned with
how early the GW networks can detect the inspiral before the merger.

@@

%Let us emphasize that this article is intended as a pedagogical introduction to gravitational-wave astronomy written at a level accessible to
%Ph.D. students, advanced undergraduates, and colleagues in astronomy and/or astrophysics who wish to learn more about the underlying physics.
%Be that as it may, even the experts might find our results motivating and exciting.
%
This article is organized as follows. Sec.~\ref{sec:BNS_inspiral} introduces the formulation for the leading-order evolution of the binary inspiral.
Sec.~\ref{sec:IFO_response} details the interferometer response to GWs based on topology.
Sec.~\ref{sec:idealizations} lists the various idealizations we employ to simplify our treatment and supplies justification for each.
Sec.~\ref{sec:GW170817} tests our model network and evolution using the parameters of GW170817 and the corresponding observations.
Sec.~\ref{sec:results} contains our main results presented as advance warning times in Tables~\ref{table:LIGO2020} and \ref{table:ET}; 
and as horizon distances and event rates in Table~\ref{table:horizon}.
In Sec.~\ref{sec:BH_NS}, we recompute the advance warning times in the case of binary black hole - neutron star inspirals to see whether or not 
future detectors can forecast a tidal disruption event, for which we summarize our findings in Table~\ref{table:ET_BH_NS}.

Throughout the text, $t$ denotes observer/detector time and $f$ denotes the GW frequency of the dominant quadrupole mode in observer/detector frame. 
%$f$ is twice the orbital (Keplerian) frequency $f_K$. 
%Relatedly, we define the following angular frequencies: $\omega = 2\pi f, \Omega=2\pi f_K$ with $\omega=2\Omega$.
We employ the $\simeq$ symbol when displaying the numerical values of quantities %with infinite decimal expansions 
which we usually truncate at four significant digits.
The $\approx$ symbol is reserved for approximations whereas $\sim$ denotes rough, order-of-magnitude equalities,
e.g., $\pi \simeq 3.142$, $\pi \approx 3$, $\pi \sim \ord(1)$. %Kepler's third law: $f_K^2\sim r^{-3}$.
Overdots denote time derivatives with respect to detector-frame time, e.g., $\dot{E} =dE/dt$, and $\propto$ denotes proportionalities. 
Unless otherwise noted, we use standard SI units.

\section{Binary neutron star inspirals}\label{sec:BNS_inspiral}
\subsection{Newtonian evolution}
For us, the starting point for the evolution of a binary under gravitational radiation reaction
%the emission of GWs 
is the expression for the power emitted in GWs given by 
the celebrated Einstein quadrupole formula \cite{1918SPAW}
%for the generated by the time-variance of the mass quadrupole moment
%
\be
\dot{E} = \f{G}{5c^5}\langle \dddot{Q}_{ij}\dddot{Q}_{ij} \rangle \label{eq:Edot_quadrupole}\, .
\ee
%
Here, $Q_{ij}$ is a trace-reversed mass quadrupole moment which
is the dominant contribution to the power emitted %in gravitational radiation
by a system with changing mass moments. Details of how to obtain this expression can be found in Ref.~\cite{Maggiore}.
There is no monopole radiation due to conservation of mass and no dipole radiation due to conservation of linear and angular momenta
(cf. Ref.~\cite{Mono}).

@@

An important quantity in GW astronomy is the chirp mass of the binary given by
%
\be
M_c = \mu^{3/5} M^{2/5} = \f{(m_1 m_2)^{3/5}}{(m_1+m_2)^{1/5}} \label{eq:chirp_mass}\, ,
\ee
%
which turns Eq.~(\ref{eq:Edot}) into
%
\be
\dot{E} = \f{32}{5}\f{c^5}{G} \left(\f{G M_c\, \omega}{2c^3}\right)^{10/3}= \f{32}{5}\f{c^5}{G} \nu^2 x^5, \label{eq:Edot2}
\ee
%
where $\omega=2\Omega$ is the angular frequency of the quadrupolar GWs, $\nu\equiv \mu/M$ is the symmetric mass ratio and $x = ({G M\, \Omega}/{c^3})^{2/3} $ is the dimensionless inverse separation parameter ubiquitous 
in post- Newtonian theory. 
As $x\propto v^2/c^2$, it is used to keep track of post-Newtonian orders and we can see from Eq.~(\ref{eq:Edot2}) that 
at the leading order, power emitted in GWs scales as $x^5$.
%hence the common statement in the literature saying that gravitational radiation starts

@@

We can also compute the number of GW cycles over the course of an inspiral. Given that the orbital period $T_K(t)$ varies over a time scale much larger than $T_K$ itself
we can write
%
\be
\mathcal{N}_\text{cyc}(f) = \int_f^{f_\text{max}} \f{f'}{\dot{f'}}\, df' =\f{1}{32\pi} \f{c^5}{(\pi G M_c)^{5/3}}\left[ f^{-5/3}-f_\text{max}^{-5/3}\right] \label{eq:Ncyc1}
\ee
%
We now introduce a cut-off for the inspiral imposed by the frequency of the innermost stable circular orbit (ISCO) in Schwarzschild spacetime
\be
\Omega_\text{ISCO} = \f{c^3}{6^{3/2} G M} \label{eq:Sch_f_isco}.
\ee
%
%
We will see in Sec.~\ref{sec:idealizations} that $f_\text{ISCO} \sim \ord(10^3)\,$Hz for a system with $M =2.8 M_\odot$. 
On the other hand, the frequencies of interest for our advance warning time estimations are $\lesssim 10\,$Hz
meaning that $f^{-5/3} \gg f_\text{ISCO}^{-5/3}$. Therefore, at the leading (Newtonian) order, the number of GW cycles can be approximated by
%
\be
\mathcal{N}_\text{cyc}(f) \approx \f{1}{32\pi} \f{c^5}{(\pi G M_c)^{5/3}}\, f^{-5/3} \,  \label{eq:Ncyc2}
\ee
%
%Restoring some familiar numbers in this expression yields
%
which can be rewritten as
\be
\mathcal{N}_\text{cyc}(f)\approx 1.605\times 10^4 \left(\f{10\,\text{Hz}}{f} \right)^{5/3}\,\left(\f{1.219 M_\odot}{M_c}\right)^{5/3}\label{eq:Ncyc3}\, .
\ee
%
This expression is quite telling: if a ground-based interferometer picks up an
inspiralling BNS at $f=10\,$Hz then there will be $\gtrsim \ord(10^4)$ GW cycles in the detector's data stream until the merger.

@@

Let us now turn our attention to the GWs generated by the inspiral.
These are tensorial perturbations propagating at the speed of light in flat (Minkowski) spacetime.
In the so-called wave zone, defined by the condition $ \lambda= c/f \ll D$, where $\lambda$ is the GW wavelength and $D$ is the distance to the GW source, 
these tensor modes satisfy the flat spacetime sourceless wave equation. %$\Box\bar{h}_{\mu\nu} = 0$.
%meaning that we are considering the solutions far from the GW source, 
In a suitable gauge, such as the commonly used transverse-traceless gauge, 
it can be shown that there exist only two physical degrees of freedom which manifest themselves as two independent polarization amplitudes $h_+$ and $h_\times$ of the GWs. %when interacting with masses away from the source.
For a binary in a circular orbit, these states read \cite{Maggiore}
%
\begin{align}
 h_+(t) &= h_c(t)\, \left(\f{1+\cos^2\iota}{2}\right)\, \cos[\Phi_\text{N}(t)],\label{eq:hplus_TD}\\
 h_\times(t) & =h_c(t)\,\cos\iota \sin[\Phi_\text{N}(t)]\label{eq:hcross_TD},
\end{align}
%
%
where
%
\be
h_c(t) =\f{4}{D}\left(\f{G M_c}{c^2}\right)^{5/3}\, \left(\f{\pi f(t)}{c}\right)^{2/3} \label{eq:h_c}
\ee
is the characteristic strain and
$\iota=\cos^{-1}(\mathbf{\hat{n}}\cdot{\mathbf{\hat{L}}}) $ is the inclination angle between the line of sight unit vector $\mathbf{\hat{n}} $ and the orbital angular momentum unit vector $\mathbf{\hat{L}}$. 
For an orbit seen edge on, $\iota = \pi/2$ giving $h_\times=0$, i.e,  a purely plus-polarized wave.
When $\iota=0$ the orbit is seen face-on and we have a circularly-polarized wave: $\langle h_+ \rangle = \langle h_\times\rangle$ where $\langle \ldots \rangle$
denote time-averages over one orbit. Note that as $f$ increases in time, so does $h(t)$, hence the characteristic ``chirping'' of GW signals.

@@

The nomenclature ``plus'' ($+$) and ``cross'' ($\times$) follows from the effects that passing GWs have on the plane transverse to their direction of propagation.
If we were to conceive of a ``toy'' detector made up of a circular configuration of test-masses lying in the $x$-$y$ plane, a GW propagating along the $z$ direction would stretch/compress this circular arrangement in a $+$ and $\times$ pattern 
with a $90^\circ$ phase difference between the two states.
This is the manifestation of the tidal strain of the GWs on the tranverse plane.
%which is why $h_{+},h_{\times}$ are often called GW strains. 
Next, let us briefly explore how an actual detector, i.e., an interferometer, responds to passing GWs.

>>>>

\begin{abstract}
We explore the intriguing possibility of employing future ground-based gravitational-wave interferometers to detect the inspiral of binary neutron stars sufficiently
early to alert electromagnetic observatories so that a gamma-ray burst (GRB) can be observed in its entirety from its very beginning.
We quantify the ability to predict a GRB by computing the time a binary neutron star (BNS) system takes to inspiral from its moment of detection to its final merger. We define the moment of detection to be the instant at which the interferometer network accumulates a signal-to-noise ratio of 15. %for the BNS inspiral.
For our computations, %of advance warning times, 
we specifically consider BNS systems at luminosity distances of (i) $D\le200\,$Mpc for the three-interferometer Advanced-LIGO-Virgo network of 2020, and (ii) $D \le 1000\,$Mpc for Einstein Telescope's B and C configurations. 
In the case of Advanced LIGO-Virgo we find that we may at best get a few minutes of warning time, thus we expect no forecast of GRBs in the 2020s. 
On the other hand, Einstein Telescope will provide us with advance warning times of more than five hours for $D \le 100\,$Mpc.
Taking one hour as a benchmark advance warning time, we obtain a corresponding horizon
distance of roughly 600 Mpc for the Einstein Telescope C configuration.
%, more specifically, greater than an hour
%for BNS inspirals closer than roughly 600\,Mpc. Taking this luminosity distance to be our horizon distance, 
Using current BNS merger event rates within this volume, we show that Einstein C will forecast $\gtrsim \ord(10^2)$ GRBs in the 2030s. %with Einstein C. %with the C configuration of Einstein Telescope.
We reapply our warning-time computation to binary black hole - neutron star inspirals and find that we expect 1 to 3 tidal disruption events 
to be forecast by the same detector.
This article is intended as a pedagogical introduction to gravitational-wave astronomy written at a level accessible to
Ph.D. students, advanced undergraduates, and colleagues in astronomy and/or astrophysics who wish to learn more about the underlying physics.
Though many of our results may be known to the experts, they might nonetheless find this article motivating and exciting.
\end{abstract}
\maketitle

\section{Introduction}\label{Sec:Intro}
The 6.95\footnote{There have thus far been six detections with $5\sigma$ statistical significance and one additional event, LVT151012 \cite{TheLIGOScientific:2016pea}, with $2\sigma \simeq 95\%$ significance hence 6.95 overall.} 
gravitational-wave merger events detected by the Advanced LIGO-Virgo network have firmly established gravitational-wave astronomy
as an observational science \cite{TheLIGOScientific:2016pea, Abbott:2017gyy, Abbott:2017oio, Abbott:2017vtc, GW170817}. 
Though the first event of 14 September 2015 (GW150914 \cite{GW150914}) will always be the ``poster-child'' of this field,
the icing on the cake was the 17 August 2017 (GW170817 \cite{GW170817}) event involving the inspiral and merger of a binary neutron star system 
promptly followed, 1.7 seconds later, by a gamma-ray burst (GRB170817A) detected by the Fermi Gamma-ray Burst Monitor \cite{Fermi} and by the International Gamma-Ray Astrophysics Laboratory \cite{Svinkin, Savchenko:2017ffs}.
The Advanced-LIGO-Virgo network's initial source localization to within $ 31\,\text{degree}^2$ in the Southern skies
enabled astronomers to locate the electromagnetic counterpart in the galaxy NGC 4993, 
first picked up by the One-Meter, Two Hemisphere team less than 11 hours after the merger \cite{Coulter2017, Coulter:2017wya}.
This subsequently launched a massive campaign of multi-messenger astronomy across the entire electromagnetic (EM) spectrum which is still ongoing
a year after the initial GRB \cite{GBM:2017lvd}.

GW170817 was initially identified by the LIGO-Hanford interferometer (H1) 
using a template bank of gravitational waveforms computed within the framework of post-Newtonian theory \cite{Buonanno:2009zt, Blanchet_LRR}.
The inspiral swept across the Advanced LIGO-Virgo (ALV) network's bandwidth from approximately $ 30\,$Hz to about $ 2\,$kHz in 57 seconds,
executing nearly $ 3000$ gravitational wave cycles and accumulating a network signal-to-noise ratio of 32.4 with a false-alarm rate of one per $8.0\times 10^4$ years \cite{GW170817}. 
The total mass of the system was inferred to be $2.74^{+0.04}_{-0.01} M_\odot$ with component masses in the range $1.17-1.60 M_\odot$, consistent
with neutron stars (assuming low spin priors, see Sec.~\ref{sec:idealizations}).
Moreover, both the gravitational wave (GW) and EM observations were consistent with a source location in the galaxy in NGC4993 
at a luminosity distance of $\sim\! 40\,$Mpc, and supported the hypothesis that GW170817 resulted from the
inspiral and merger of two neutron stars, causing the prompt short-hard gamma-ray burst GRB170817A. %as 
This BNS-GRB connection was first proposed in Ref.~\cite{Eichler:1989ve} and the events of 17 August 2017 are
the first astrophysical evidence for it.

A week after GW170817, the ALV network was taken offline for upgrades for the third observation run (O3) 
scheduled to start in early 2019 with roughly twice the sensitivity of O2.
In addition, the Japanese cryogenic interferometer KAGRA (\cite{KAGRA, KAGRA2}) will start its test runs in 2018-2019 and reach its design sensitivity circa 2021-2022 \cite{Akutsu:2017thy, Aasi:2013wya}. By 2025, LIGO-India detector will join the global interferometer network hence improving both the overall sensitivity and the sky-localization capabilities \cite{Aasi:2013wya}.
Around this time, the construction of the first third-generation ground-based interferometer, Einstein Telescope, will begin in Europe \cite{ET_doc}. 
Meanwhile, the US is currently considering a mid-2020s update to LIGO called Voyager \cite{LIGO_Voy} and the construction of a more ambitious 40\,km long interferometer dubbed Cosmic Explorer \cite{CE}.
All of these near-future additions to the GW-detector network beg a simple question: can we detect the inspiral of a binary neutron star system
early enough to witness a gamma-ray burst as it happens? In other words, can we forecast GRBs using GW detections?

The short answer is ``yes''. The long answer is still ``yes'', but depends on many factors such as source distance and sky position, detector sensitivity and orientation. One also needs to account for the fact that GRBs are thought to be collimated emissions with narrow outburst angles \cite{Kumar:2014upa}.
As such, most GRBs that occur in the universe are electromagnetically undetectable. Fortunately, this was not the case with GRB170817A, 
but even for a reasonable GRB jet opening angle of $\sim 10^\circ$, we expect to observe at best $\sim 10\%$ of neutron star mergers as GRBs \cite{Patricelli:2016bkt}. 
However, it is possible that some of the near-Earth pointing short hard GRBs may be observed as long-duration GRBs or X-ray flashes thus increasing this percentage \cite{Bucciantini:2011kx}.

@@

Here, we are interested  in exploring the predictive power of near-future ground-based GW detectors. 
%thus leave the task of reducing the latency to the experts.
More specifically, we wish to find out how much early warning future GW detectors can provide us before the merger/GRB.
Ref.~\cite{Cannon:2011vi} approached this from an algorithmic perspective by developing computationally inexpensive methods to detect inspiral
signals in GW data to provide early-warning triggers \emph{before} the merger (also see Ref.~\cite{Knowles:2018hqq} for a very recent improvement to what is currently being
used in the ALV pipeline).
We instead focus on the forecasting capabilities of the ground-based interferometers at their \emph{full design} sensitivity by computing the time interval
between the merger and the instant of detection defined in terms of a certain detection criterion which we describe below.
To this end, we consider the ALV network of 2020s composed of three L-shaped interferometers operating at their design sensitivities
and Einstein Telescope of 2030s made up of V-shaped interferometers.
Ref.~\cite{Nissanke:2012dj} investigated the capabilities of three to five-interferometer detector networks of 2020s in terms of their horizon distances
to inspiralling BNSs and their sky-localization of these sources. Our work here is somewhat complementary in the sense that we are mostly concerned with
how early the GW networks can detect the inspiral before the merger.

@@

%Let us emphasize that this article is intended as a pedagogical introduction to gravitational-wave astronomy written at a level accessible to
%Ph.D. students, advanced undergraduates, and colleagues in astronomy and/or astrophysics who wish to learn more about the underlying physics.
%Be that as it may, even the experts might find our results motivating and exciting.
%
This article is organized as follows. Sec.~\ref{sec:BNS_inspiral} introduces the formulation for the leading-order evolution of the binary inspiral.
Sec.~\ref{sec:IFO_response} details the interferometer response to GWs based on interferometer topology.
Sec.~\ref{sec:idealizations} lists the various idealizations we employ to simplify our treatment and supplies justification for each.
Sec.~\ref{sec:GW170817} tests our model network and evolution using the parameters of GW170817 and the corresponding observations.
Sec.~\ref{sec:results} contains our main results presented as advance warning times in Tables~\ref{table:LIGO2020} and \ref{table:ET}; 
and as horizon distances and event rates in Table~\ref{table:horizon}.
In Sec.~\ref{sec:BH_NS}, we recompute the advance warning times in the case of binary black hole - neutron star inspirals to see whether or not 
future detectors can forecast tidal disruption events, for which we summarize our findings in Table~\ref{table:ET_BH_NS}.

Throughout the text, $t$ denotes observer/detector time and $f$ denotes the \emph{GW frequency} of the dominant quadrupole mode in observer/detector frame. 
%$f$ is twice the orbital (Keplerian) frequency $f_K$. 
%Relatedly, we define the following angular frequencies: $\omega = 2\pi f, \Omega=2\pi f_K$ with $\omega=2\Omega$.
We employ the $\simeq$ symbol when displaying the numerical values of quantities %with infinite decimal expansions 
which we usually truncate at four significant digits.
The $\approx$ symbol is reserved for approximations whereas $\sim$ denotes rough, order-of-magnitude equalities,
e.g., $\pi \simeq 3.142$, $\pi \approx 3$, $\pi \sim \ord(1)$. %Kepler's third law: $f_K^2\sim r^{-3}$.
Overdots denote time derivatives with respect to detector-frame time, e.g., $\dot{E} =dE/dt$, and $\propto$ denotes proportionalities. 
Unless otherwise noted, we use standard SI units.

\section{Binary neutron star inspirals}\label{sec:BNS_inspiral}
\subsection{Newtonian evolution}
For us, the starting point for the evolution of a binary under gravitational radiation reaction
%the emission of GWs 
is the expression for the power emitted in GWs given by 
the celebrated Einstein quadrupole formula \cite{1918SPAW}
%for the generated by the time-variance of the mass quadrupole moment
%
\be
\dot{E} = \f{G}{5c^5}\langle \dddot{Q}_{ij}\dddot{Q}_{ij} \rangle \label{eq:Edot_quadrupole}\, .
\ee
%
Here, $Q_{ij}$ is a trace-reversed mass quadrupole moment which
gives the dominant contribution to the power emitted %in gravitational radiation
by a system with changing mass moments. Details of how to obtain this expression can be found in Ref.~\cite{Maggiore}.
There is no monopole radiation due to conservation of mass and no dipole radiation due to conservation of linear and angular momenta
(cf. Ref.~\cite{Mono}).

@@

An important quantity in GW astronomy is the chirp mass of the binary given by
%
\be
M_c = \mu^{3/5} M^{2/5} = \f{(m_1 m_2)^{3/5}}{(m_1+m_2)^{1/5}} \label{eq:chirp_mass}\, ,
\ee
%
which turns Eq.~(\ref{eq:Edot}) into
%
\be
\dot{E} = \f{32}{5}\f{c^5}{G} \left(\f{G M_c\, \omega}{2c^3}\right)^{10/3}= \f{32}{5}\f{c^5}{G} \nu^2 x^5, \label{eq:Edot2}
\ee
%
where $\omega=2\Omega$ is the angular frequency of the quadrupolar GWs, $\nu\equiv \mu/M$ is the symmetric mass ratio and $x = ({G M\, \Omega}/{c^3})^{2/3} $ is the dimensionless inverse separation parameter ubiquitous 
in post- Newtonian theory. 
As $x\propto v^2/c^2$, it is used to keep track of post-Newtonian orders and we can see from Eq.~(\ref{eq:Edot2})
that, at the leading order, power emitted in GWs scales as $x^5$.
%hence the common statement in the literature saying that gravitational radiation starts

@@

We can also compute the number of GW cycles over the course of an inspiral. Given that the orbital period $T_K(t)$ varies over a time scale much larger than itself
we can write
%
\be
\mathcal{N}_\text{cyc}(f) = \int_f^{f_\text{max}} \f{f'}{\dot{f'}}\, df' =\f{1}{32\pi} \f{c^5}{(\pi G M_c)^{5/3}}\left[ f^{-5/3}-f_\text{max}^{-5/3}\right] \label{eq:Ncyc1}
\ee
%
We now introduce a cut-off for the inspiral imposed by the frequency of the innermost stable circular orbit (ISCO) in Schwarzschild spacetime
\be
\Omega_\text{ISCO} = \f{c^3}{6^{3/2} G M} \label{eq:Sch_f_isco}.
\ee
%
%
We will see in Sec.~\ref{sec:idealizations} that $f_\text{ISCO} \sim \ord(10^3)\,$Hz for a system with $M =2.8 M_\odot$. 
On the other hand, the frequencies of interest for our advance warning time estimations are $\lesssim 10\,$Hz
meaning that $f^{-5/3} \gg f_\text{ISCO}^{-5/3}$. Therefore, at the leading (Newtonian) order, the number of GW cycles can be approximated by
%
\be
\mathcal{N}_\text{cyc}(f) \approx \f{1}{32\pi} \f{c^5}{(\pi G M_c)^{5/3}}\, f^{-5/3} \,  \label{eq:Ncyc2}
\ee
%
%Restoring some familiar numbers in this expression yields
%
which can be rewritten as
\be
\mathcal{N}_\text{cyc}(f)\approx 1.605\times 10^4 \left(\f{10\,\text{Hz}}{f} \right)^{5/3}\,\left(\f{1.219 M_\odot}{M_c}\right)^{5/3}\label{eq:Ncyc3}\, .
\ee
%
This expression is quite telling: if a ground-based interferometer picks up an
inspiralling BNS at $f=10\,$Hz then there will be $\gtrsim \ord(10^4)$ GW cycles in the detector's data stream until the merger.

@@

Let us now turn our attention to the GWs generated by the inspiral.
These are tensorial perturbations propagating at the speed of light in flat (Minkowski) spacetime.
In the so-called wave zone, defined by the condition $ \lambda= c/f \ll D$, where $\lambda$ is the GW wavelength and $D$ is the distance to the GW source, 
these tensor modes satisfy the flat spacetime sourceless wave equation. %$\Box\bar{h}_{\mu\nu} = 0$.
%meaning that we are considering the solutions far from the GW source, 
%
It can further be shown that there exist only two radiative degrees of freedom which physically manifest themselves as two independent polarization amplitudes $h_+$ and $h_\times$ of the GWs 
This fact is most easily demonstrated in the so-called transverse-traceless gauge (cf. Sec.~2.1 of Ref.~\cite{Flanagan:2005yc}). %when interacting with masses away from the source.

%In a suitable gauge, such as the commonly used transverse-traceless gauge, 
%In the ``lab'' frame, usually taken to be the transverse-traceless gauge, 
The two polarization states generated by a binary in a circular orbit read \cite{Maggiore}
%
\begin{align}
 h_+(t) &= h_c(t)\, \left(\f{1+\cos^2\iota}{2}\right)\, \cos[\Phi_\text{N}(t)],\label{eq:hplus_TD}\\
 h_\times(t) & =h_c(t)\,\cos\iota \sin[\Phi_\text{N}(t)]\label{eq:hcross_TD},
\end{align}
%
%
where
%
\be
h_c(t) =\f{4}{D}\left(\f{G M_c}{c^2}\right)^{5/3}\, \left(\f{\pi f(t)}{c}\right)^{2/3} \label{eq:h_c}
\ee
is the characteristic strain and
$\iota=\cos^{-1}(\mathbf{\hat{n}}\cdot{\mathbf{\hat{L}}}) $ is the inclination angle between the line of sight unit vector $\mathbf{\hat{n}} $ and the orbital angular momentum unit vector $\mathbf{\hat{L}}$. 
For an orbit seen edge on, $\iota = \pi/2$ giving $h_\times=0$, i.e,  a purely plus-polarized wave.
When $\iota=0$ the orbit is seen face-on and we have a circularly-polarized wave: $\langle h_+ \rangle = \langle h_\times\rangle$ where $\langle \ldots \rangle$
denote time-averages over one orbit. Note that as $f$ increases in time, so does $h(t)$, hence the characteristic ``chirping'' of GW signals.

@@

The nomenclature ``plus'' ($+$) and ``cross'' ($\times$) follows from the effects that passing GWs have on the plane transverse to their direction of propagation.
If we were to conceive of a ``toy'' detector made up of a circularly arranged test-masses lying in the $x$-$y$ plane, a GW propagating along the $z$ direction would stretch/compress this circular arrangement in a $+$ and $\times$ pattern 
with a $90^\circ$ phase difference between the two states.
This is the manifestation of the tidal strain of the GWs on the tranverse plane.
%which is why $h_{+},h_{\times}$ are often called GW strains. 
Next, let us briefly explore how an actual detector, i.e., an interferometer, responds to passing GWs.

\section{Interferometer response to gravitational waves}\label{sec:IFO_response}
\subsection{LIGO-Virgo configuration: L-shaped topology}\label{sec:LIGO_topo}
We start by first considering the L-shaped interferometer topology dating back to Michelson \cite{Michelson:1887zz}.
Although GWs are described by propagating tensor modes, an interferometer (IFO) can only measure a scalar quantity
known as the response function (or GW strain) which is a linear combination of the polarizations given by
%
\be
h(t) = F_+(\theta,\phi,\psi) \,h_+(t)+F_\times(\theta,\phi,\psi)\, h_\times(t) , \label{eq:detector_strain}
\ee
%
where
%%
%
\begin{align}
 F_+ &=\f{1}{2}\left(1+\cos^2\theta\right)\cos2\phi\cos 2\psi-\cos\theta \sin2\phi \sin2\psi \label{eq:F_plus},\\
 F_\times &=\f{1}{2}\left(1+\cos^2\theta\right)\cos2\phi\sin2\psi+\cos\theta \sin2\phi \cos2\psi \label{eq:F_cross}
\end{align}
%
%%
%$F_+, F_\times$ 
are the antenna pattern functions of the detectors; $\theta,\phi$ are the sky coordinates; and $\psi$ is
the angle of orientation of a detector's frame with respect to the source frame. %(there is no need for $\psi$ in the case of a single detector).
%In principle, $\psi$ can be aligned with the polarization of the incoming GWs 
%
We can not apriori know $\theta, \phi,\psi$, however, we can extract their values with some limited precision over the course of an inspiral.
The longer the inspiral stays in the detector bandwidth, the better we can estimate these angles.
%Thus, BNS inspirals which would sweep across the detection band over thousands of cycles are preferable. 
In the the case of GW170817 which swept the ALV network over roughly 3000 GW cycles, 
the initial source sky localization was 31\,deg$^2$ (later reduced to 28\,deg$^2$) using data from three IFOs \cite{GW170817}. 
For our estimations, we employ suitable averages of these angles as explained below.
For details on how to obtain Eqs.~(\ref{eq:F_plus}, \ref{eq:F_cross}) see, e.g., Sec.~4.2.1 of Ref.~\cite{SchutzLRR}.
